An Overview on Temporal Databases.\\ New Approaches and Proposals.

### 2012-03-09 book/sources/timeDomain.tex
A temporary version of the proces of implementing Guy's remarks

\textcolor{red}{CHECK}
Restrictions on time range may be performed on the base of the following two concepts~\cite{Jensen94thetsql2}:
\begin{itemize}
\item
\textbf{Time boundedness}: Time can be bounded orthogonally in the past and in the future.
\item
\textbf{Time distance}: As a metric, time has a distance function which presents four properties:
\begin{enumerate}
\item
Distance is non-negative.
\item
The distance between two different elements is not equal to zero.
\item
The distance from $\alpha$ to $\beta$ is the same as that between $\beta$ and $\alpha$.
\item
Triangle inequality: The distance from $\alpha$ to $\gamma$ is equal or greater than the distance from $\alpha$ to $\beta$ plus the distance from $\beta$ to $\gamma$.
\end{enumerate}
\end{itemize}
\textcolor{red}{END CHECK}

\textcolor{red}{MOVE TO TEMPORAL DATABASE PART}
On the other hand, time may be~\cite{Jensen94thetsql2} \textbf{relative} or \textbf{absolute} (in other words, \textbf{anchored} or \textbf{unanchored}). Sometimes, what one considers \emph{absolute time} is not as definite as one would hope since our concept of absolute time is based on another time reference (for example, January 1st of year one). Relative time has a direction, differently from distance. It is also possible to use negative references so as to represent relative time (e.g. -3 days = three days ago).
\textcolor{red}{END MOVE}

Subsection \ref{subsec:granularity} gives a short survey on the concept of temporal granularities.

@@

It is also possible to describe the relation between different granularities. This is called a casting function:
\begin{svgraybox}
\begin{definition}
\textbf{Casting function}~\cite{Lin97}\\
\label{def:casting-fucntion}
A casting function $cast$ maps granules from a granularity $G$ to granules from a different granularity $H$.
\begin{equation}
\label{eq:cast-function}
\mbox{cast} \left( i_G,G,H \right) = j_H \nonumber
\end{equation}
\end{definition}
\end{svgraybox}

This notation indicates that the function $cast$ associates a granule $i_G$ in $G$ to a corresponding granule $j_H$ in $H$. Two kinds of granularity-to-granularity mappings can now be discerned: an \emph{upwards mapping} is a mapping from a granularity $G$ to a coarser granularity $H$, whereas a downwards mapping is a mapping from a granularity $K$ to a finner granularity $L$. Different granularity-to-granularity mappings between several granularities can be represented using a granularity graph, which is a directed graph indicating the mapping conversions.

 Mappings between two granularities may be classified as \emph{regular mappings}, \emph{irregular mappings} or \emph{congruent mappings}.
\begin{itemize}
\item
\emph{Regular mapping}: A regular mapping is a granularity-to-granularity mapping where the mapping function between the two granularities is calculated by means of multiplications and/or divisions and (maybe) an anchor adjustment. E.g. the mapping between hours and minutes is calculated using a multiplication by $60$. 
\item
\emph{Irregular mapping}: An irregular mapping is a granularity-to-granularity mapping where the mapping function can not be calculated by means of multiplications and/or divisions. E.g. the mapping between months and days is dependent on the exact month or day. 
\item
\emph{Congruent mapping}: A congruent mapping is a granularity-to-granularity mapping where the two granularities in the mapping have the same granules but a different anchor. E.g. the mapping between the days (Gregorian calendar days) and the academic days. 
\end{itemize}

\begin{example}
Consider a systems that models both Gregorian Calendar dates as well as Academic dates. The chronons granularity in this system are milliseconds. Figure \ref{fig:granularity-graph-example} shows the complete granularity graph example. The transitions between Chronons and Seconds and between Seconds and Minutes and so on is called a \emph{regular mapping}. The transition between Days and Months is not defined by a multiplication or division, this is called a \emph{irregular mapping}. In the graph, these \emph{irregular mappings} are noted with a bold arrow. Finally, the transition between a Gregorian Calendar Day and an Academic Day is called a \emph{Congruent mapping}: two granularities with the same granules but with different anchors. Both concern the same days, but the academic year starts on October 1st, whereas the Gregorian calendar year starts on January 1st. Congruent mappings are visualised as straight lines without arrow heads.

@@

Next, in subsection \ref{subsec:timedomain-calendar}, some basic concepts and issues concerning calendars are explained.

\subsection{Calendars}
\label{subsec:timedomain-calendar}
%maybe introduce previously the concept of time units, linear hierarchy and time point.

This subsection will concern \emph{calendars}. A calendar provides some sort of human interpretation of time. As such, calendars provide meaning or interpretation to temporal values where this meaning or interpretation is relevant or useful to the user. More precisely, calendars determine the mapping between human-meaningful time values and an underlying time model~\cite{Dyreson1994}. More formally, a calendar is an organization of different granularities (e.g. granularities day, month and year) in a hierarchy. In order to formally define what a calendar is, some other concepts should first be introduced~\cite{Kraus1997}:

@@

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Linear Hierarchy of Granularities}\\
A \emph{\textbf{linear hierarchy of granularities}} is a finite set of granularities with a linear order (denoted $\sqsubset$). The most coarse granularity is called \emph{the top of the hierarchy} and must be an infinite set while all other granularities must be finite sets.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

An example of a linear hierarchy is the set $\left\{ \text{days, months, years} \right\}$, where days $\sqsubset$ months and months $\sqsubset$ years. With this, a \emph{linear calendar} can be defined.

\begin{svgraybox}
\vspace{-10pt}
\begin{definition}\textbf{Linear Calendar}\\
A \emph{\textbf{linear calendar}} consist of a linear hierarchy of granularities and a validity predicate which specifies if a granule is valid in the calendar.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

An example of a linear calendar could be the linear hierarchy $\left\{ \text{days, months, years} \right\}$, where days $\sqsubset$ months and months $\sqsubset$ years, together with a validity predicate that excludes time indications like `32 April 2012' or `30 February 2012'.

@@

The next section briefly introduces temporal relationships.

>>>>

%\textcolor{red}{CHECK}
Restrictions on time range may exist, as time may be bounded orthogonally in the past and in the future~\cite{Jensen94thetsql2}.
%\begin{itemize}
%\item
%\textbf{Time boundedness}: Time can be bounded orthogonally in the past and in the future.
%\item
%\textbf{Time distance}: As a metric, time has a distance function which presents four properties:
%\begin{enumerate}
%\item
%Distance is non-negative.
%\item
%The distance between two different elements is not equal to zero.
%\item
%The distance from $\alpha$ to $\beta$ is the same as that between $\beta$ and $\alpha$.
%\item
%Triangle inequality: The distance from $\alpha$ to $\gamma$ is equal or greater than the distance from $\alpha$ to $\beta$ plus the distance from $\beta$ to $\gamma$.
%\end{enumerate}
%\end{itemize}
%\textcolor{red}{END CHECK}

%\textcolor{red}{MOVE TO TEMPORAL DATABASE PART}
%On the other hand, time may be~\cite{Jensen94thetsql2} \textbf{relative} or \textbf{absolute} (in other words, \textbf{anchored} or \textbf{unanchored}). Sometimes, what one considers \emph{absolute time} is not as definite as one would hope since our concept of absolute time is based on another time reference (for example, January 1st of year one). Relative time has a direction, differently from distance. It is also possible to use negative references so as to represent relative time (e.g. -3 days = three days ago).
%\textcolor{red}{END MOVE}

%Subsection \ref{subsec:granularity} gives a short survey on the concept of temporal granularities.

@@

It is also possible to describe the relation between different granularities. This is called a casting function:
\begin{svgraybox}
\vspace{-10pt}
\begin{definition}
\textbf{Casting function}~\cite{Lin97}\\
\label{def:casting-fucntion}
Consider two different granularities $G$ and $H$. A granularity-to-granularity \emph{\textbf{casting function}} $cast$ is then a function mapping granules from $G$ to granules from $H$: 
\begin{align}
\label{eq:cast-function}
%\mbox{cast} \left( i_G,G,H \right) = j_H \nonumber
cast & : G \times \mathbb{G} \times \mathbb{G} \rightarrow H \nonumber\\
		 & : (i_{G}, G, H) \rightarrow j_{H} \nonumber
\end{align}
where $i_{G} \in G$ and $j_{H} \in H$ and where $\mathbb{G}$ denotes the set of all granularities.
\end{definition}
\vspace{-10pt}
\end{svgraybox}

Thus, the function $cast$ associates a granule $i_G$ in $G$ to a corresponding granule $j_H$ in $H$. Two kinds of granularity-to-granularity mappings can now be discerned: an \emph{upwards mapping} is a mapping from a granularity $G$ to a coarser granularity $H$, whereas a downwards mapping is a mapping from a granularity $K$ to a finner granularity $L$. Orthogonal to this classification, mappings between two granularities may be classified as \emph{regular mappings}, \emph{irregular mappings} or \emph{congruent mappings}~\cite{Lin97},~\cite{DyresonSnodgrass1994}.
\begin{itemize}
\item
\emph{Regular mapping}: A regular mapping is a granularity-to-granularity mapping where the mapping function value is calculated by means of multiplications and/or divisions and (maybe) an anchor adjustment. E.g. the mapping value of the mapping between hours and minutes is calculated using a multiplication by $60$. 
\item
\emph{Irregular mapping}: An irregular mapping is a granularity-to-granularity mapping where the mapping function value can not be calculated by means of multiplications and/or divisions. E.g. the mapping value of the mapping between months and days is dependent on the exact month or day. 
\item
\emph{Congruent mapping}: A congruent mapping is a granularity-to-granularity mapping where the two granularities involved in the mapping have the same granules but a different anchor. E.g. the mapping between the days (Gregorian calendar days) and the academic days is a congruent mapping.
\end{itemize}

Different granularity-to-granularity mappings between several granularities can be represented using a granularity graph, which is a directed graph indicating the mapping conversions. The above is illustrated in the following example.

\begin{example}
Consider a system that models both Gregorian calendar dates as well as academic calendar dates. In this system, the chronons granularity is a set of milliseconds. Figure \ref{fig:granularity-graph-example} shows the complete granularity graph corresponding to this example. The transitions between the chronons granularity and the seconds granularity and between the seconds granularity and the minutes granularity are examples of regular mappings. They are represented by normal arrows in the visualisation of the graph. The transition between the days granularity and the months granularity is an example of an irregular mapping. In the graph visualisation, irregular mappings are represented by a bold arrow. Finally, the transition between the Gregorian calendar day granularity and the academic day granularity is an example of a congruent mapping. Both concern the same days, but the academic year starts on October 1st, whereas the Gregorian calendar year starts on January 1st. In the graph visualisation, congruent mappings are visualised as straight lines without arrow heads.

@@

%Next, in subsection \ref{subsec:timedomain-calendar}, some basic concepts and issues concerning calendars are explained.

%\subsection{Calendars}
%\label{subsec:timedomain-calendar}
%maybe introduce previously the concept of time units, linear hierarchy and time point.

%This subsection will concern \emph{calendars}. A calendar provides some sort of human interpretation of time. As such, calendars provide meaning or interpretation to temporal values where this meaning or interpretation is relevant or useful to the user. More precisely, calendars determine the mapping between human-meaningful time values and an underlying time model~\cite{Dyreson1994}. More formally, a calendar is an organization of different granularities (e.g. granularities day, month and year) in a hierarchy. In order to formally define what a calendar is, some other concepts should first be introduced~\cite{Kraus1997}:

@@

%\begin{svgraybox}
%\vspace{-10pt}
%\begin{definition}\textbf{Linear Hierarchy of Granularities}\\
%A \emph{\textbf{linear hierarchy of granularities}} is a finite set of granularities with a linear order (denoted $\sqsubset$). The most coarse granularity is called \emph{the top of the hierarchy} and must be an infinite set while all other granularities must be finite sets.
%\end{definition}
%\vspace{-10pt}
%\end{svgraybox}

%An example of a linear hierarchy is the set $\left\{ \text{days, months, years} \right\}$, where days $\sqsubset$ months and months $\sqsubset$ years. With this, a \emph{linear calendar} can be defined.

%\begin{svgraybox}
%\vspace{-10pt}
%\begin{definition}\textbf{Linear Calendar}\\
%A \emph{\textbf{linear calendar}} consist of a linear hierarchy of granularities and a validity predicate which specifies if a granule is valid in the calendar.
%\end{definition}
%\vspace{-10pt}
%\end{svgraybox}

%An example of a linear calendar could be the linear hierarchy $\left\{ \text{days, months, years} \right\}$, where days $\sqsubset$ months and months $\sqsubset$ years, together with a validity predicate that excludes time indications like `32 April 2012' or `30 February 2012'.

@@

%The next section briefly introduces temporal relationships.
